---
title: "trajSurvR Demo"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{trajSurvR Demo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", warning = FALSE, message = FALSE)
```

## Overview

This vignette demonstrates the core workflow wrapped from the `raw/*.Rmd` analyses:

1. QC  
2. Differential analysis  
3. Correlation study  
4. Trajectory modeling  
5. Time-to-event analysis (baseline Cox + LASSO Cox + time-dependent Cox)  
6. Event prediction (logistic regression)  
7. Time-of-onset prediction (regression with baseline)  
8. Feature selection (LASSO + forward greedy)

```{r load-pkg}
if (!requireNamespace("trajSurvR", quietly = TRUE)) {
  # Development mode (run inside repository before installation)
  r_dir <- if (dir.exists("R")) "R" else "../R"
  r_files <- list.files(r_dir, full.names = TRUE, pattern = "\\.R$")
  invisible(lapply(r_files, source))
} else {
  library(trajSurvR)
}
```

```{r load-data}
example_df <- simulate_trajsurv_data(n_subjects = 180, n_visits = 4, seed = 2026, generic_feature_names = TRUE)
head(example_df[, c("SampleID", "UIDx", "Group", "CollAge", "delta_001", "protein_001")])
```

## 1) QC

```{r qc}
# Build a synthetic long-format table to demonstrate QC function
qc_long <- dplyr::tibble(
  SampleID = rep(example_df$SampleID, 3),
  Assay = rep(c("protein_001", "protein_002", "protein_003"), each = nrow(example_df)),
  NPX = c(example_df$protein_001, example_df$protein_002, example_df$protein_003),
  SampleQC = "PASS",
  AssayQC = "PASS"
)

qc_res <- qc_olink_data(qc_long, missing_rate_threshold = 0.20)
summarize_qc(qc_res)
```

## 2) Differential analysis

```{r differential}
proteins_npx <- c("protein_001", "protein_002", "protein_003")
diff_res <- fit_differential_mixed_models(
  analysis_df = example_df,
  proteins = proteins_npx,
  group_col = "Group",
  constant_covariates = c("Sex", "GenoGroup"),
  time_varying_covariates = c("CollAge", "Weight", "Height"),
  id_col = "UIDx",
  case_level = "Convert",
  ref_level = "Control",
  min_rows = 10
)
diff_res
```

```{r differential-plot, fig.width=7, fig.height=5}
plot_differential_volcano(diff_res)
```

## 3) Correlation study

```{r correlation}
corr_res <- compute_correlation_study(example_df, proteins = proteins_npx, method = "pearson")
corr_res$correlation_matrix
```

```{r correlation-plot, fig.width=6, fig.height=5}
plot_correlation_heatmap(corr_res)
```

## 4) Trajectory modeling

```{r trajectory}
delta_wide <- example_df |>
  dplyr::select(SampleID, delta_001, delta_002, delta_003)

visit_design <- example_df |>
  dplyr::filter(Group == "Convert") |>
  dplyr::select(SampleID, UIDx, YrSinceOs, Sex, GenoGroup, CollAge, Weight, Height)

traj_res <- run_trajectory_batch(
  delta_wide = dplyr::filter(delta_wide, SampleID %in% visit_design$SampleID),
  visit_design = visit_design,
  proteins = c("delta_001", "delta_002"),
  id_col = "UIDx",
  time_col = "YrSinceOs",
  constant_covariates = c("Sex", "GenoGroup"),
  time_varying_covariates = c("CollAge", "Weight", "Height"),
  min_obs = 5
)
traj_res$summary
```

```{r trajectory-plot, fig.width=7, fig.height=5}
if ("delta_001" %in% names(traj_res$models)) {
  plot_trajectory_fit(traj_res$models[["delta_001"]], time_col = "YrSinceOs")
}
```

## 5) Time-to-event analysis (multi-model)

```{r tte}
protein_surv <- example_df |>
  dplyr::select(SampleID, delta_001, delta_002, delta_003)

surv_dat <- build_baseline_survival_data(
  visit_info = example_df,
  protein_wide = protein_surv,
  sample_id_col = "SampleID",
  id_col = "UIDx",
  group_col = "Group",
  converter_label = "Convert",
  pre_converter_label = "Potential Convert",
  onset_col = "YrSinceOs",
  censor_col = "YrSinceCen",
  constant_covariates = c("Sex", "GenoGroup"),
  time_varying_covariates = c("CollAge", "Weight", "Height")
)

cox_uni <- fit_baseline_cox(
  dat = surv_dat,
  proteins = c("delta_001", "delta_002", "delta_003"),
  constant_covariates = c("Sex", "GenoGroup"),
  time_varying_covariates = c("CollAge", "Weight", "Height"),
  min_rows = 8,
  min_events = 2
)
cox_uni
```

```{r cox-plot, fig.width=7, fig.height=5}
plot_cox_forest(cox_uni, top_n = 10)
```

```{r lasso-cox}
if (nrow(surv_dat) >= 30) {
  lasso_cox <- fit_lasso_cox(
    dat = surv_dat,
    proteins = c("delta_001", "delta_002", "delta_003"),
    constant_covariates = c("Sex", "GenoGroup"),
    time_varying_covariates = c("CollAge", "Weight", "Height"),
    nfolds = 3
  )
  lasso_cox$selected_min
} else {
  cat("Example dataset is small; skipping LASSO Cox fit in this demo.\n")
}
```

```{r td-cox}
td_base <- example_df |>
  dplyr::select(UIDx, time_since_baseline, stop_time, event, CollAge, Weight, Height, Sex, GenoGroup, delta_001, delta_002, delta_003)

intervals <- make_counting_process_data(
  visit_df = td_base,
  id_col = "UIDx",
  visit_time_col = "time_since_baseline",
  stop_time_col = "stop_time",
  event_col = "event",
  keep_visit_cols = c("CollAge", "Weight", "Height", "Sex", "GenoGroup", "delta_001", "delta_002", "delta_003")
)

td_data <- intervals

td_res <- fit_time_dependent_cox(
  td_data = td_data,
  proteins = c("delta_001", "delta_002"),
  constant_covariates = c("Sex", "GenoGroup"),
  time_varying_covariates = c("CollAge", "Weight", "Height"),
  min_subjects = 5,
  min_events = 2
)
td_res
```

```{r td-cox-plot, fig.width=7, fig.height=4.5}
plot_cox_forest(td_res, top_n = 10)
```

## 6) Event prediction (logistic regression)

```{r logistic}
pred_dat <- make_event_outcome(example_df, horizon_years = 2, group_col = "Group", onset_col = "YrSinceOs") |>
  dplyr::filter(!is.na(CASE))

features <- c("delta_001", "delta_002", "CollAge")
features <- c(features, "Weight", "Height", "Sex")
auc_cv <- cv_logistic_auc(pred_dat, feature_cols = features, outcome_col = "CASE", n_folds = 3)
auc_cv
```

## 7) Time-of-onset prediction (regression with baseline)

```{r onset-regression}
onset_dat <- example_df |>
  dplyr::filter(Group %in% c("Potential Convert", "Convert")) |>
  dplyr::mutate(Y = ifelse(is.na(Y), 0, Y))

onset_fit <- fit_onset_time_regression(
  data = onset_dat,
  outcome_col = "Y",
  feature_cols = c("delta_001", "delta_002"),
  constant_covariates = c("Sex", "GenoGroup"),
  time_varying_covariates = c("CollAge", "Weight", "Height"),
  id_col = "UIDx"
)
onset_fit$metrics
```

```{r onset-plot, fig.width=6, fig.height=5}
plot_onset_predictions(onset_fit)
```

## 8) Feature selection (LASSO + forward greedy)

```{r fs}
class_counts <- table(pred_dat$CASE)
if (length(class_counts) == 2 && all(class_counts >= 2)) {
  lasso_event <- fit_event_prediction_lasso(
    data = pred_dat,
    outcome_col = "CASE",
    feature_cols = c("delta_001", "delta_002", "delta_003", "CollAge", "Weight", "Height"),
    nfolds = 3
  )
  lasso_event$selected_min
} else {
  cat("Example dataset has too few cases/controls for logistic LASSO; skipping.\n")
}

greedy <- forward_greedy_selection(
  data = pred_dat,
  candidate_features = c("delta_001", "delta_002", "delta_003", "Weight", "Height"),
  base_features = c("CollAge", "Sex"),
  evaluator = function(dat, feats) cv_logistic_auc(dat, feature_cols = feats, outcome_col = "CASE", n_folds = 3),
  min_improvement = 0
)
greedy$history
```

```{r greedy-plot, fig.width=7, fig.height=5}
if (nrow(greedy$history) > 0) {
  plot_greedy_history(greedy)
}
```
