---
title: "trajSurvR: End-to-End Workflow Demo"
subtitle: "Longitudinal Trajectory and Survival Analysis for Proteomic Data"
author: "Ximing Ran"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
    fig_height: 4.5
    fig_width: 7
    latex_engine: xelatex
    highlight: tango
geometry: margin=2.5cm
fontsize: 11pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  warning   = FALSE,
  message   = FALSE,
  fig.align = "center",
  out.width = "90%"
)
```

\newpage

# Overview

This document demonstrates the core analytical workflow implemented in the
**trajSurvR** package.  All analyses use a fully synthetic dataset generated by
`simulate_trajsurv_data()`, so no real patient data are required to run this
demo.

**Workflow modules covered:**

1. Simulate data
2. Quality control (QC)
3. Differential expression (linear mixed models)
4. Correlation study
5. Trajectory modelling (GAM/GAMM)
6. Time-to-event analysis — baseline Cox, LASSO Cox, time-dependent Cox
7. Event prediction (logistic regression + CV AUC)
8. Time-of-onset prediction (regression with LOOCV)
9. Feature selection — LASSO and forward greedy search

**Installation:**

```r
remotes::install_github("ranxm2/trajSurvR")
```

---

# Load Package

```{r load-pkg}
if (!requireNamespace("trajSurvR", quietly = TRUE)) {
  r_dir <- if (dir.exists("../R")) "../R" else "R"
  invisible(lapply(list.files(r_dir, full.names = TRUE, pattern = "\\.R$"), source))
} else {
  library(trajSurvR)
}
suppressPackageStartupMessages(library(dplyr))
```

\newpage

# Simulate Data

`simulate_trajsurv_data()` generates a realistic longitudinal cohort with three
subject groups: **Control**, **Potential Convert**, and **Convert**.  Each
subject has multiple visits, protein measurements (`protein_001`–`003`), change
scores (`delta_001`–`003`), time-varying covariates (age, weight, height), and a
survival outcome.

```{r simulate}
dat <- simulate_trajsurv_data(
  n_subjects = 180,
  n_visits   = 4,
  seed       = 2026,
  generic_feature_names = TRUE
)

cat("Dimensions:", nrow(dat), "rows x", ncol(dat), "columns\n")
cat("Groups:\n")
print(table(dat$Group))
cat("\nColumn names:\n")
print(colnames(dat))
```

```{r show-head, echo=FALSE}
knitr::kable(
  head(dat[, c("SampleID", "UIDx", "Group", "CollAge", "Weight",
               "delta_001", "protein_001", "event")]),
  caption = "First six rows of the simulated dataset.",
  booktabs = TRUE
)
```

\newpage

# Quality Control

`qc_olink_data()` expects Olink-style long-format data.  We reshape the wide
simulated data into long format as a demonstration.

```{r qc}
qc_long <- tibble(
  SampleID = rep(dat$SampleID, 3),
  Assay    = rep(c("protein_001", "protein_002", "protein_003"),
                 each = nrow(dat)),
  NPX      = c(dat$protein_001, dat$protein_002, dat$protein_003),
  SampleQC = "PASS",
  AssayQC  = "PASS"
)

qc_res <- qc_olink_data(qc_long, missing_rate_threshold = 0.20)
```

```{r qc-table, echo=FALSE}
knitr::kable(
  summarize_qc(qc_res),
  caption = "QC stage-wise sample and protein counts.",
  booktabs = TRUE
)
```

\newpage

# Differential Expression

One linear mixed model per protein, group effect adjusted for sex, genotype,
age, weight, and height.  A random intercept is fitted per subject.

```{r differential}
proteins_npx <- c("protein_001", "protein_002", "protein_003")

diff_res <- fit_differential_mixed_models(
  analysis_df            = dat,
  proteins               = proteins_npx,
  group_col              = "Group",
  constant_covariates    = c("Sex", "GenoGroup"),
  time_varying_covariates = c("CollAge", "Weight", "Height"),
  id_col                 = "UIDx",
  case_level             = "Convert",
  ref_level              = "Control",
  min_rows               = 10
)
```

```{r diff-table, echo=FALSE}
knitr::kable(
  diff_res[, c("Protein", "N", "coef", "se", "p_value", "fdr", "direction",
               "significant")],
  digits  = 4,
  caption = "Mixed-model differential analysis results.",
  booktabs = TRUE
)
```

```{r diff-volcano, fig.cap="Volcano plot: effect size vs −log10(p). Orange points are FDR < 0.05."}
plot_differential_volcano(diff_res)
```

\newpage

# Correlation Study

```{r correlation}
corr_res <- compute_correlation_study(dat, proteins = proteins_npx,
                                      method = "pearson")
```

```{r corr-table, echo=FALSE}
knitr::kable(
  round(corr_res$correlation_matrix, 3),
  caption = "Pearson correlation matrix.",
  booktabs = TRUE
)
```

```{r corr-heat, fig.cap="Heatmap of pairwise Pearson correlations."}
plot_correlation_heatmap(corr_res)
```

\newpage

# Trajectory Modelling

GAM/GAMM is fitted to the `delta` (change score) of each protein over time
relative to disease onset, in the converter group only.

```{r trajectory-data}
delta_wide   <- select(dat, SampleID, delta_001, delta_002, delta_003)
visit_design <- dat |>
  filter(Group == "Convert") |>
  select(SampleID, UIDx, YrSinceOs, Sex, GenoGroup, CollAge, Weight, Height)
```

```{r trajectory}
traj_res <- run_trajectory_batch(
  delta_wide   = filter(delta_wide, SampleID %in% visit_design$SampleID),
  visit_design = visit_design,
  proteins     = c("delta_001", "delta_002"),
  id_col       = "UIDx",
  time_col     = "YrSinceOs",
  constant_covariates     = c("Sex", "GenoGroup"),
  time_varying_covariates = c("CollAge", "Weight", "Height"),
  min_obs = 5
)
```

```{r traj-table, echo=FALSE}
knitr::kable(
  traj_res$summary[, c("Protein", "N_obs", "N_subjects", "p_value", "fdr",
                       "first_sig_time")],
  digits  = 4,
  caption = "Trajectory batch summary.",
  booktabs = TRUE
)
```

```{r traj-plot, fig.cap="Fitted trajectory for delta\\_001 (observed points, smooth + 95\\% CI)."}
if ("delta_001" %in% names(traj_res$models)) {
  plot_trajectory_fit(traj_res$models[["delta_001"]], time_col = "YrSinceOs")
}
```

\newpage

# Time-to-Event Analysis

## Build Baseline Survival Dataset

```{r tte-build}
protein_surv <- select(dat, SampleID, delta_001, delta_002, delta_003)

surv_dat <- build_baseline_survival_data(
  visit_info              = dat,
  protein_wide            = protein_surv,
  sample_id_col           = "SampleID",
  id_col                  = "UIDx",
  group_col               = "Group",
  converter_label         = "Convert",
  pre_converter_label     = "Potential Convert",
  onset_col               = "YrSinceOs",
  censor_col              = "YrSinceCen",
  constant_covariates     = c("Sex", "GenoGroup"),
  time_varying_covariates = c("CollAge", "Weight", "Height")
)
cat("Baseline survival dataset:", nrow(surv_dat), "subjects,",
    sum(surv_dat$event), "events\n")
```

## Univariate Baseline Cox

```{r baseline-cox}
cox_uni <- fit_baseline_cox(
  dat                     = surv_dat,
  proteins                = c("delta_001", "delta_002", "delta_003"),
  constant_covariates     = c("Sex", "GenoGroup"),
  time_varying_covariates = c("CollAge", "Weight", "Height"),
  min_rows   = 8,
  min_events = 2
)
```

```{r cox-table, echo=FALSE}
knitr::kable(
  cox_uni[, c("Protein", "N", "N_events", "coef", "HR", "p_value", "fdr",
              "direction")],
  digits  = 4,
  caption = "Univariate baseline Cox results.",
  booktabs = TRUE
)
```

```{r cox-forest, fig.cap="Forest plot of univariate baseline Cox hazard ratios."}
plot_cox_forest(cox_uni, top_n = 10)
```

## LASSO Cox

```{r lasso-cox}
if (nrow(surv_dat) >= 30) {
  lasso_cox <- fit_lasso_cox(
    dat                     = surv_dat,
    proteins                = c("delta_001", "delta_002", "delta_003"),
    constant_covariates     = c("Sex", "GenoGroup"),
    time_varying_covariates = c("CollAge", "Weight", "Height"),
    nfolds = 3
  )
  cat("LASSO Cox selected (lambda.min):", paste(lasso_cox$selected_min, collapse = ", "), "\n")
  cat("LASSO Cox selected (lambda.1se):", paste(lasso_cox$selected_1se, collapse = ", "), "\n")
}
```

## Time-Dependent Cox

```{r td-cox}
td_base <- select(dat, UIDx, time_since_baseline, stop_time, event,
                  CollAge, Weight, Height, Sex, GenoGroup,
                  delta_001, delta_002, delta_003)

intervals <- make_counting_process_data(
  visit_df       = td_base,
  id_col         = "UIDx",
  visit_time_col = "time_since_baseline",
  stop_time_col  = "stop_time",
  event_col      = "event",
  keep_visit_cols = c("CollAge", "Weight", "Height", "Sex", "GenoGroup",
                      "delta_001", "delta_002", "delta_003")
)

td_res <- fit_time_dependent_cox(
  td_data                 = intervals,
  proteins                = c("delta_001", "delta_002"),
  constant_covariates     = c("Sex", "GenoGroup"),
  time_varying_covariates = c("CollAge", "Weight", "Height"),
  min_subjects = 5,
  min_events   = 2
)
```

```{r td-table, echo=FALSE}
knitr::kable(
  td_res[, c("Protein", "N_subjects", "N_events", "coef", "HR", "p_value",
             "fdr")],
  digits  = 4,
  caption = "Time-dependent Cox results.",
  booktabs = TRUE
)
```

```{r td-forest, fig.cap="Forest plot of time-dependent Cox hazard ratios."}
plot_cox_forest(td_res, top_n = 10)
```

\newpage

# Event Prediction

Binary outcome: did the subject convert within the next 2 years?

```{r logistic}
pred_dat <- make_event_outcome(
  dat,
  horizon_years    = 2,
  group_col        = "Group",
  onset_col        = "YrSinceOs"
) |> filter(!is.na(CASE))

features <- c("delta_001", "delta_002", "CollAge", "Weight", "Height", "Sex")

auc_cv <- cv_logistic_auc(
  pred_dat,
  feature_cols = features,
  outcome_col  = "CASE",
  n_folds      = 3
)
cat("Cross-validated AUROC:", round(auc_cv, 4), "\n")
```

\newpage

# Time-of-Onset Prediction

Linear regression with LOOCV to predict years to disease onset.

```{r onset-regression}
onset_dat <- dat |>
  filter(Group %in% c("Potential Convert", "Convert")) |>
  mutate(Y = ifelse(is.na(Y), 0, Y))

onset_fit <- fit_onset_time_regression(
  data                    = onset_dat,
  outcome_col             = "Y",
  feature_cols            = c("delta_001", "delta_002"),
  constant_covariates     = c("Sex", "GenoGroup"),
  time_varying_covariates = c("CollAge", "Weight", "Height"),
  id_col                  = "UIDx"
)
```

```{r onset-metrics, echo=FALSE}
knitr::kable(
  onset_fit$metrics,
  digits  = 3,
  caption = "LOOCV prediction metrics for onset-time regression.",
  booktabs = TRUE
)
```

```{r onset-plot, fig.cap="Observed vs LOOCV-predicted onset time."}
plot_onset_predictions(onset_fit)
```

\newpage

# Feature Selection

## LASSO Logistic

```{r lasso-event}
class_counts <- table(pred_dat$CASE)
if (length(class_counts) == 2 && all(class_counts >= 2)) {
  lasso_event <- fit_event_prediction_lasso(
    data         = pred_dat,
    outcome_col  = "CASE",
    feature_cols = c("delta_001", "delta_002", "delta_003",
                     "CollAge", "Weight", "Height"),
    nfolds = 3
  )
  cat("LASSO logistic selected (lambda.min):",
      paste(lasso_event$selected_min, collapse = ", "), "\n")
}
```

## Forward Greedy Selection

```{r greedy}
greedy <- forward_greedy_selection(
  data               = pred_dat,
  candidate_features = c("delta_001", "delta_002", "delta_003",
                         "Weight", "Height"),
  base_features      = c("CollAge", "Sex"),
  evaluator          = function(d, f)
    cv_logistic_auc(d, feature_cols = f, outcome_col = "CASE", n_folds = 3),
  min_improvement    = 0
)
```

```{r greedy-table, echo=FALSE}
knitr::kable(
  greedy$history,
  digits  = 4,
  caption = "Forward greedy selection history.",
  booktabs = TRUE
)
```

```{r greedy-plot, fig.cap="AUC gain at each step of forward greedy selection."}
if (nrow(greedy$history) > 0) {
  plot_greedy_history(greedy)
}
```

\newpage

# Session Info

```{r session}
sessionInfo()
```
